---
title: "Search, Annotation, and R Script Example"
author: "Robert Fromont"
date: "4 February 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Search, Annotation, and R Script Example}
  %\usepackage[UTF-8]{inputenc}
---


[LaBB-CAT](https://labbcat.canterbury.ac.nz) is a browser-based linguistic annotation store that stores audio or video recordings, 
text transcripts, and other annotations. The *nzilbb.labbcat* R package provides access to linguistic data stored in LaBB-CAT 
servers, allowing tokens and their annotations to be identified and extracted, along with media data, and acoustic measurements.

This vignette shows how to:

1. identify a set of phone tokens in specific contexts, 
2. extract annotation labels for the tokens and their speakers,
2. download audio files for all tokens,
3. execute a custom R script to extract acoustic measurements from the audio files.

## The effects of morphology and demographics on 'sibilance' of /s/ 

In particular, we are interested in whether the /s/ in a *mis* in a prefixed word like *mistimes* is pronounced differently from the *mis* in an unprefixed word like *mistakes*, and whether the age or gender of the speaker makes any difference.

First the *nzilbb.labbcat* package must be loaded, and the LaBB-CAT corpus is specified:

```{r nzilbb.labbcat, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
require(nzilbb.labbcat)
labbcat.url <- "https://labbcat.canterbury.ac.nz/demo/"
```

```{r credentials, eval=TRUE, echo=FALSE, render=FALSE}
credentialError <- labbcatCredentials(labbcat.url, "demo", "demo")
```

Now we search our LaBB-CAT corpus for aligned words that start "dis" or "mis".

```{r searches, warning=FALSE}
matches <- getMatches(labbcat.url, list(orthography = "[dm]is.+"), aligned=T, no.progress = T)

print(head(matches)[, c("Transcript", "Target.transcript", "Target.transcript.start")]) ## TODO transcript->word
```

In order to be able to analyse the /s/ segments, we need it's start and end time, so we can extract a matching sound sample. We identify the first three *segments* annotations of each token, the last of which will be the /s/ segment we're interested in.

```{r segments}
segments <- getMatchAlignments(
  labbcat.url, matches$MatchId, c("segments"), annotationsPerLayer=3) ## TODO segments->segment

matches <- cbind(matches, segments)
print(head(matches)[, c(
  "Text", "segments.3", "segments.3.start", "segments.3.end")]) ## TODO segments->segment
```

We want to check whether the 'sibilance' of /s/ varies depending on whether it's in a morphological prefix or not, so we need the morphological parses of each of the word tokens. We also extract some demographic information about the speakers.

```{r morphology-and-demographics}
morphology.demographics <- getMatchLabels(
  labbcat.url, matches$MatchId, c("morphology", "participant_age_category", "participant_gender"))

matches <- cbind(matches, morphology.demographics)
print(head(matches)[, c(
  "Text", "morphology", "participant_age_category", "participant_gender")])
```

We want to perform acoustic analysis, so we need the wav sample for each match, which can be extracted from LaBB-CAT as as a mono 22kHz files using `getSoundFragments`.

```{r process.row, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
## define subdirectory to save the files in
subdir <- "s-tokens"

## get segment sound files from LaBB-CAT
wav.files <- getSoundFragments(
  labbcat.url, matches$Transcript, matches$segments.3.start, matches$segments.3.end, 22050, 
  path=subdir, no.progress = T) ## segments->segment

print(head(wav.files))
```

Now that we have tokens of /s/, we want to use R to perform acoustic analysis of the sound of each token, and save the resulting acoustic measurements in the data frame with the results.

Koenig *et al.* (2013) propose an 8-factor multitaper spectrum calculated over 25 ms portions of the segment, i.e. 8 independent estimates of the spectrum in a single time window are calculated and averaged. We will use a custom R script (adapted from a script written by Patrick Reidy, who has kindly given permission for its use in this example) to provide a function called `process.wav` which takes a .wav file for a given token and returns the Koenig *et al.* (2013) measures.

This function is defined in [hayhawkins-multitaper.R](https://github.com/nzilbb/labbcat-R/blob/master/vignettes/examples/hayhawkins-multitaper.R), which must be loaded.

```{r hayhawkins-multitaper, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
source('hayhawkins-multitaper.R')
```

The `process.wav` function processes a single sound file, so we `apply` it to all our /s/ tokens, and add the resulting acoustic measures to our *matches* data frame.

```{r apply-wav.files, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
s.measures.lists <- lapply(wav.files, process.wav)

require(data.table) ## we need this package in order to use rbindlist below
matches <- cbind(matches, as.data.frame(rbindlist(s.measures.lists)))
```

Now that the sound files we downloaded have been processed, we don't need them any more, so we tidily delete them to save disk space.

```{r tidy-up, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
removed <- file.remove(wav.files)
removed <- file.remove(subdir)
```

The `process.wav` function produces 12 acoustic measures, three of which are printed below by way of example:

- *freqM* - peak frequency in the 'medium' range (3kHz to 7kHz)
- *freqH* - peak frequency in the 'high' range (7kHz to the Nyquist frequency)
- *CoG* - the centre of gravity of the whole spectrum

```{r example-measures, eval=TRUE, echo=TRUE, render=FALSE, warning=FALSE}
print(head(matches)[, c("Number", "Text", "segments.3.start", "freqM", "freqH", "CoG")])
```

The dataset now includes acoustic measurements allowing study the 'sibiliance' of /s/, in relation to morphology and speaker demographics.
